---
title: "Final_project_script"
author: "Maryela C"
format: html
editor: visual
---

## PART 1: Data description 

Growth data come from the Forest Macrosystems Network (FMN) project, belonging to the forests of Niwot Ridge, CO, USA and Barro Colorado Island, Panama; while drought data were downloaded from SPEI Global Drought Monitor.

Load data

```{r}
#Packages
library(readxl)

#data
NWT<- read_excel("C:/Users/Usuario/Downloads/ECOL_596_Finalproject_Maryela/01_Cleaning_dataframes/NWT_growthSPEI_cleaned.xlsx")


BCI<- read_excel("C:/Users/Usuario/Downloads/ECOL_596_Finalproject_Maryela/01_Cleaning_dataframes/BCI_growthSPEI_cleaned.xlsx")

```

## PART 2: Scientific goal 

The research question of this project is "Are drought seasons reflected on the growth rates?"

## PART 3: Statistical analysis 

To test if drought seasons are affecting the growth rates, I conducted a Linear Mixed-Effect Model (LME) per site, where SPEI was use as the fixed effect, and a random effect was included for individual trees and years. Then, I used a LME, where SPEI where a fixed effect and the Forest type was an interaction term, with the same random effects.

0.  Load necessary library

```{r}
library(Matrix)
library(lme4)
library(lmerTest)
library(forestmangr)
library(dplyr)
```

Pre-statistical analysis settings:

1.  Check for NA or empty strings on the df

    ```{r}
    #NIWOT RIDGE FOREST
    NWT$dbh <- as.numeric(NWT$dbh)
    head(NWT$dbh)

    #SAME FOR BARRO COLORADO ISLAND
    BCI$dbh <- as.numeric(BCI$dbh)
    head(BCI)
    ```

2.  Used the tree_summarise function from the "forestmangr" package to calculate the equivalent diameter of trees with more than one trunk.

```{r}
#For Niwot Ridge Forest
NWT<-tree_summarise(NWT, #my DF
                    "dbh", #column with size measurements
                    tree="main", #tree id column
                    .groups = c("plot","date.year","site","line")) #parameters to subdivide the estimated dbh


#Same for Barro Colorado Island Forest
BCI<-tree_summarise(BCI, #my DF
                    "dbh", #column with size measurements
                    tree="main", #tree id column
                    .groups = c("plot","date.year","site","line")) #parameters to subdivide the estimated dbh



```

2.  Then, calculated the "growth rate". A simple way I thought I could get the growth rate, is by subtracting the value of dbh from the immediately preceding year. There are specific formulas to calculate this, but for practical purposes, I decided to use this method.

```{r}
#For NWT
NWT_rate <- NWT %>%
  group_by(main) %>%
  arrange(main, date.year) %>%  # Ensure it's sorted by year within each tree
  mutate(growth_rate = c(NA, diff(dbh))) %>%
  ungroup()  # Remove grouping after calculation



#Same for BCI
BCI_rate <- BCI %>%
  group_by(main) %>%
  arrange(main, date.year) %>%  # Ensure it's sorted by year within each tree
  mutate(growth_rate = c(NA, diff(dbh))) %>%
  ungroup()  # Remove grouping after calculation

```

3.  As a tree cannot "decrease", I omitted all negative "growth_rate" values, as this is probably due to measurement error.

    ```{r}
    #FOR NWT
    NWT_rate_c <- NWT_rate %>%
      filter(growth_rate >= 0)


    #SAME FOR BCI
    BCI_rate_c <- BCI_rate %>%
      filter(growth_rate >= 0)

    ```

### LME for Niwot Ridge

Model 1 was the simplest and the one that best fit the data (lower REML). This suggests a weak positive relationship between drought and growth rate (Estimate=0.036 , p-value=0.069).

Notice that I also tried others models, such as:

-   Model 2: growth_rate \~ SPEI + (1 \| species_binomial)

    -   The REML (31674.1) was higher than Model 1, but the difrection of the effect was the same: positive relationship between SPEI and growth rate.

-   Model 3: growth_rate \~ SPEI + (1 \| species_binomial) + ( 1 \| year)

    -   REML (31673.1) higher than Model 1 but similar to Model 2.

-   Model 4: growth_rate \~ SPEI + ( 1 \| tree tag) + (1 \| year)

    -   REML (30871.6) lower tan Model 1, but p-value suggested no evidence for a relationship between drought and growth.

```{r}
#Here I'm showing MODEL 1
NWT_lme<- lmer(growth_rate ~ SPEI_3 + (1 | main) , data = NWT_rate_c)

summary(NWT_lme)

```

### LME for Barro Colorado Island

```{r}

```

Both

## PART 4: Results interpretation
