---
title: "Cleaning_DFs"
author: "mcr"
format: html
editor: visual
---

## Steps followed for cleaning the data

upload data and packages

```{r}
library(readr)
library(dplyr)
library(readxl)
library(openxlsx)

##NIWOT
NWT_SPEI_3 <- read_csv("01_Cleaning_dataframes/NWT_SPEI_3.csv")


NWT_data <- read_excel("01_Cleaning_dataframes/NWT_data.xlsx")

```

First, calculated the average SPEI per year, per site.

```{r}
# Extract the year from the DATA column
NWT_SPEI_3$year <- substr(NWT_SPEI_3$DATA, 5, 6)

# Convert the year to a full year format (e.g., "51" to "1951")
NWT_SPEI_3$year <- ifelse(as.numeric(NWT_SPEI_3$year) < 50, 
                          paste0("20", NWT_SPEI_3$year),  # Assume years after 2000 for <50
                          paste0("19", NWT_SPEI_3$year))  # Assume years before 2000 for >=50


# Calculate the mean SPEI for each year
NWT_annual_spei <- aggregate(SPEI_3 ~ year, data = NWT_SPEI_3, FUN = mean)

```

Then, added the mean SPEI to my growth dataset

```{r}
colnames(NWT_annual_spei)[colnames(NWT_annual_spei) == "year"] <- "date.year"


# Merge the datasets by the "year" column
NWT_data <- merge(NWT_data, NWT_annual_spei, by = "date.year", all.x = TRUE)
View(NWT_data)
```

Then, saved the data frames by site

```{r}
write.xlsx(NWT_data, "NWT_growthSPEI.xlsx")



```

Selecting the data

First, I removed the rows where the dbh value is less than the dbh value of the previous year, based on the idea that "a tree can not decrease", and if it decreased may actually be because it was measured wrong.

```{r}
NWT_growthSPEI <- read_excel("01_Cleaning_dataframes/NWT_growthSPEI.xlsx")

#Order the data by year
NWT_growthSPEI <- NWT_growthSPEI[order(NWT_growthSPEI$date.year), ]

#created a new column with the previous dbh to easily compare
NWT_growthSPEI$prev_dbh <- c(NA, NWT_growthSPEI$dbh[-nrow(NWT_growthSPEI)])

#remove rows of trees that "decreased"
NWT_growthSPEI <- NWT_growthSPEI[NWT_growthSPEI$dbh >= NWT_growthSPEI$prev_dbh, ]

#delete the column with the previous year data
NWT_growthSPEI$prev_dbh <- NULL
```

I then omitted dgh values, as younger trees may have a different response/behaviour to drought events.

```{r}
##TO OMIT DGH VALUES
NWT_growthSPEI <- NWT_growthSPEI[!is.na(NWT_growthSPEI$dbh), ]


# deleted dgh column
NWT_growthSPEI <- subset(NWT_growthSPEI, select = -dgh)

```

Finally, to make it easier to use other functions in the future, I filled in the "main" column with the same value as "enq" if it was empty (in my DF, when "main" has a number", it indicates the tag of the main individual, and the "enq" value is from a tag for another arm o trunk of the same individual)

```{r}
NWT_growthSPEI$main[NWT_growthSPEI$main %in% c("NA", "na")] <- NA

NWT_growthSPEI$main <- trimws(NWT_growthSPEI$main) # Remove spaces
NWT_growthSPEI$main[NWT_growthSPEI$main == ""] <- NA # Replace empty strings with NA


NWT_growthSPEI$main <- ifelse(is.na(NWT_growthSPEI$main), 
                              NWT_growthSPEI$enq, 
                              NWT_growthSPEI$main)

```

Second round of cleaning

Check for NA or empty strings on the DF

```{r}
   #NIWOT RIDGE FOREST
   NWT_growthSPEI$dbh <- as.numeric(NWT_growthSPEI$dbh)
   head(NWT_growthSPEI)
   
# Remove rows with NA in the species_binomial column
NWT_growthSPEI <- na.omit(NWT_growthSPEI, cols = "species_binomial")
```

For practical purposes, also decided to omit all "unidentified" trees

```{r}

NWT_growthSPEI_clean <- NWT_growthSPEI %>%
  filter(!(species_binomial %in% c("unidentified_NWT", NA, "")))
```

Saved the cleaned DFs

```{r}
write.xlsx(NWT_growthSPEI_clean, "NWT_c.xlsx")
```
